\LARGE \bf Bringing robotics into formal education \\ using the Thymio open source hardware robot

### 2016-10-19 paper_ram.tex
finalization

\author{Francesco Mondada$^{1}$, Michael Bonani$^{2}$, Fanny Riedo$^{2}$, Manon Briod$^{1}$, \\
L\'ea Pereyre$^{1}$, Philippe R\'etornaz$^{1}$ and St\'ephane Magnenat$^{1}$% <-this % stops a space
\thanks{*This research was supported by the Swiss National Center of Competence in Research ``Robotics'' (Thymio robot development and deployment), by GebertRuf Stiftung (design of accessories), by the Swiss NSF project CRAGP2 151543 ``Robotics in schools'', and by the EU-FP7 project ASSISIbf, no. 601074 (survey on open hardware).
Many thank to Luc Bergeron and his team at \'ecal.ch for the industrial design of Thymio; Didier Roy, David Sherman and all the INRIA team for their contributions and diffusion in France; Gordana Gerber for the educational material; all the Mobsya team for the effort in production and sale; and Fabian Hahn for the port of Blockly.}% <-this % stops a space``anyone''
\thanks{$^{1}$Francesco Mondada, Manon Briod, L\'ea Pereyre, Philippe R\'etornaz and St\'ephane Magnenat are with the Laboratoire de Syst\`emes Robotiques,
        Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Switzerland
        {\tt\small firstname.lastname@epfl.ch}}%
\thanks{$^{2}$Michael Bonani and Fanny Riedo are with Mobsya Association, Renens, Switzerland
        {\tt\small michael.bonani@mobsya.org} and {\tt\small fanny.riedo@mobsya.org}}%
}

@@

Mobile robots are valuable tools for education because of both the enthusiasm they raise and the multidisciplinary nature of their technology.
They give access to a wide range of fields, such as complex mechanics, sensors, wireless transmission, mathematics, computer science, etc.
As mobile robots sense the environment and take actions based on their perception, they seem to display \emph{intentions of their own}~\cite{Mondada2001}.
%This fascinates the users of the robot and creates feelings of accomplishment and power when these users engage in creative activities with the robots.
This impression of intelligent technology, the permeating presence of robots in science fiction and their projected use in our society give a sense of touching the future.
%Children can spend hours looking at a robot interacting with its environment and adults can approach robots as a new hobby that replaces radio and electronics kits.
%The emotional potential of mobile robots and their varied technology makes them a potential ideal tool for education.

Despite their potential as educational tools, robots are still not as widespread in schools as they could be. Among the possible reasons, we believe that the following five are playing a key role:
\begin{enumerate}
\item Although there are many research projects developing innovative and interesting educational robots, few reach a sufficient maturity to become distributed and accessible to schools. 
\item A versatile robot performing interesting behaviors is a complex piece of technology and therefore expensive. 
%Indeed, for having an educational value and providing an interesting level of interaction, a robot must embed a wide set of sensors and actuators.
%Existing platforms with these features cost several hundred Euros.
This prevents most schools, which have a limited budget for equipment, to acquire educational robots.
\item Introducing robotic tools into teaching activities requires investment in time and training for the teachers~\cite{CERI2008}.
%In Europe, despite a trend to increase the use of technological tools in the learning process, teachers are still insufficiently trained and are reluctant to introduce these tools in their teaching activity~\cite{CERI2008}.
Therefore, to be accepted by teachers, robots must be both accessible with minimal effort and accompanied by well-prepared educational material shared among colleagues. 
%Moreover, infrastructure should be provided to allow the sharing of educational material among teachers.
\item Robot construction, use, and programming is often perceived as a boyish activity in our society~\cite{leonard2009lego,nourbakhsh2009robot}.
% For instance, several authors report that girls have such a perception of the LEGO Minstorms~\cite{leonard2009lego,nourbakhsh2009robot}.
This strongly limits the potential of robots as general-purpose educational tools, especially in schools.
\item Finally, many teachers are reluctant to follow volatile trends, especially if these are based on purely commercial arguments.
Teachers prefer to invest in stable tools, and this is the opposite as the trend of current consumer technology. %, which is often very volatile and based on continuous renewal rather than on durability.
\end{enumerate}

Open source hardware projects can address several of these issues in a different way than closed-source purely commercial products. 
By \textit{open source hardware} we mean, following the definition of the Open Source Hardware Association\footnote{\url{http://www.oshwa.org/definition/}}, ``hardware whose design is made publicly available so that anyone can study, modify, distribute, make, and sell the design or hardware based on that design''.
In this paper, we show that this concept, implemented in the Thymio project and gathering a community of users, developers and manufacturers, brings a strong added value to the robot and to the educational methods.
In addition, we compare our experience with other robotics open source hardware projects not focused on education, and highlight challenges and opportunities specific to education.

@@

There is a large number of publications presenting educational robots, from low-cost systems targeting Africa~\cite{Rubenstein2015,Gyebi2015} to extremely sophisticated humanoids~\cite{Hood2015,Mazzoni2016}. 
Among those, only a handful are commercially available, limiting their validation by educational scientists, who are typically not roboticists.
This results in 90\,\% of publications about validation of educational results made on LEGO\textsuperscript{\textregistered} Mindstorms\textsuperscript{\textregistered}~\cite{benitti2012exploring}, a widely available commercial product.
The latest version, EV3\footnote{\url{http://mindstorms.lego.com}}, is expensive ($\approx$ 400\,\$) but offers a wide range of possibilities, especially at the mechanical level using LEGO\textsuperscript{\textregistered} bricks and at the software level with its graphical programming environment. 
%NOTE: SM: commented WeDo to save space and streamline story
%At a lower price, LEGO\textsuperscript{\textregistered} proposes the WeDo\textsuperscript{\textregistered} system ($\approx$ 80\,\$).
%The WeDo\textsuperscript{\textregistered} is similar to EV3 but only reads one sensor and controls one actuator and must always be connected to a desktop computer.
%In comparison, the BeeBot robot\footnote{\url{http://www.beebot.org.uk}} is a cheap ($\approx$ 90\,\$) mobile system.
%However, it is not a real robot but rather an automaton: There is no perception of the environment, just a good encoder for precise movements and a simple programming interface based on buttons giving directions for the movement.
Among the other players on the market, one can cite the Edison robot\footnote{\url{https://meetedison.com/}}, extremely low cost (49\,\$) and compatible with LEGO\textsuperscript{\textregistered} bricks. The low price has pushed to drastic design choices; small size, weak motors, few sensors, communication with the computer by audio jack and standard AAA batteries not included in the price.
These choices strongly limits its possible usage.
%Among widely used robots, none is part of an open source hardware project.

Among the robots available on the market, only few are open source and are used in schools:
Scribbler2, produced and sold by Parallax\footnote{\url{http://www.parallax.com}} ($\approx$ 180\,\$), is a 188\,mm large robot, designed to run on the ground and equipped with few light sensors, one distance sensor, two ground sensors and few \textsc{led} displays.
It runs on standard AA batteries and has a hacker port for interfacing electronic extensions.
It is programmable with a graphical or a textual code interface.  
The main weakness of Scribbler2 is its limited number of sensors and the limited compatibility with other systems. 
Moreover, there seems to be no active community around its development.
The e-puck~\cite{mondada2009puck} robot is a robot targeting university-level education.
Well equipped with sensors and actuators, modular and compact, it can be programmed with industry-standard environments.
Several simulators allow to run experiments of high complexity. 
%There is an active community of users and developers around it.
Its main weakness is its high price ($\approx$ 870\,\$).
Finch\footnote{\url{http://www.finchrobot.com/}} ($\approx$ 99\,\$) is a very simple robot that has been designed around a wired connection to the computer. 
This connection reduces electronics requirements, such as batteries or wireless communication, and allows to implement the control entirely on the computer.
This results in a very broad set of possible programming languages available, which is the real force of this robot. 
However, the cable does not allow real autonomy and mobility.
Finally, the mBot\footnote{\url{http://www.makeblock.cc/mbot/}} is a mobile platform based on an Arduino board.
Its electronics is simple and inexpensive and the robot only features a couple of sensors, which allows to drastically reduce its price ($\approx$ 75\,\$), but also limits the perception possibilities and therefore the span of use.

With respect to these robots, Thymio has a compact size (110\,mm), many interaction possibilities, an affordable price and a large set of sensors.
%Its design makes the best use of the recent cheap components coming from the explosive growth of the mobile devices market. 
To the best of our knowledge, beside Thymio there are no educational products providing a similar integration of sensors and actuators at a price below 130\,\$.

Teachers, from primary school up to high school, are one of the main target user group of the Thymio project.
They decide which tools are used in their class and are key people in the eduction ecosystem.
For teachers, the motivation to use robotic tools depends on many factors~\cite{chevalier2016}.
Among them, the availability of material and training plays a key role.
%The BeeBot mentioned previously, because of its radical simplicity, is one of the tools that needs the least training. 
%Moreover its use in learning activities is based on a simple grid on a printed map on the ground, allowing fast and easy integration in existing lessons.
%These are probably two important factors for its success in schools~\cite{2008highfield,2008demichele} and the availability of a rich set of educational material\footnote{\url{https://www.learningplace.com.au/deliver/content.asp?pid=38840} and \\ \url{http://beebots.skola.edu.mt/}}.
%However, the total lack of sensors reduces the BeeBot to an automaton, easy to use but limited to activities implementable by sequences of motions, without any closed-loop control that would allow teaching computational thinking or robotics concepts.
The development of educational material and courses to train the teachers requires a huge effort, based on a good mix between robotics and educational skills. 
Moreover, educational material varies from school to school, as requirements are very dependent on local educational programs and languages. 
%For example, in Switzerland, every of the 26 states composing the country has its own school program and its own language among the four national ones. 
%Educational material and the related training must therefore be adapted to each situation.
Even for large companies like LEGO\textsuperscript{\textregistered}, such a development is not sustainable. 
To solve this problem, a crowd-sourcing approach can be used.
An active community of users can contribute to the development of the material in a distributed manner, adapting the material to the local situation.
LEGO\textsuperscript{\textregistered} itself is moving toward this direction by promoting communities of users~\cite{Hienerth2014} with a user-producer interaction close to those of open hardware projects.
An open source community regrouping developers, manufacturers and end users is therefore a very interesting model to address the distributed development and sharing of educational material, and the diffusion of training sessions.
%This model is an extension to education of what is already practiced in projects such as Arduino~\cite{Jamieson2010}.
In this paper, we study a case of implementation of this model.

@@

\begin{figure}
   \centering
	\includegraphics[width=.8\columnwidth]{figures/front_2016}
	\vskip 1.0em
	\includegraphics[width=\columnwidth]{figures/wirelessThymioVecto}
	\caption{The Thymio robot and its main components for the Wireless and the \textsc{usb}-connected versions.}
	\label{fig:schematique}
\end{figure}

We designed the Thymio robot along seven main axes: a low price to address a larger number of users; a feature set that suits both genders and multiple ages from young children to adults; a mechanical design that promotes creativity; a combination of sensors, actuators, and programming features that facilitates learning: a set of ready-to-use programs to get fast access to robotic behaviors; an accessible programming environment; and an open source community contributing to the design and the dissemination.
The result is a miniature (110\,mm in side) differential-wheeled robot suited for use on a desktop (Fig.~\ref{fig:schematique}, top).
The robot is robust enough to be mishandled by children, as it can fall from a table without breaking itself.
It features a translucent white hull and a wide range of sensors and actuators (Fig.~\ref{fig:schematique}, bottom).
The robot embeds a battery, rechargeable by \textsc{usb}, that provides an autonomy of about 3 to 5 hours.
More details on the robot and the previous research results can be found in~\cite{RiedoPhD, magnenat2014}.

@@

Price is key in the adoption of robots by schools~\cite{kradolfer2014sociological}.
Therefore the design of Thymio had to target a low production cost, while including a broad range of functionalities to enable flexibility in its use. 
As in this type of robot the main cost comes from electronics and sensors~\cite{thesis_michael}, we focused on low-cost sensors that allow rich interaction possibilities, with both the environment and the user.

The resulting Thymio robot possesses a large number of simple sensors:
Seven horizontal proximity sensors, two ground sensors, a three-axis accelerometer, a thermistor, and a microphone.
Five capacitive touch buttons organized as a direction pad form an intuitive user interface.
Compared to physical buttons, these simplify the plastic hull of the robot and make it more robust.
A remote control receiver provides additional distant buttons.
%A microphone can record sounds and detect claps. 
%Finally, a thermistor measures temperature.
Most of these devices cost less then 0.2\,\$, the most expensive being the accelerometer with a cost of about 0.8\,\$, which is an acceptable price given the possibilities it brings to the robot.
%The omnipresence of this type of sensors in smartphones and compact cameras has drastically lowered their prices over the last decade.
We also chose low-cost toy motors and control them in speed (max. 13\,cm/s) measuring the back-electromotive force and therefore removing the need of additional encoders.

%However, choosing the optimal combination of sensors and processor is a difficult problem, as hardware interfaces are limited on a microcontroller and therefore compromises must be made.
We evaluated several microcontrollers and chose the PIC24F from Microchip because it integrates a \textsc{usb} interface and can drive capacitive touch buttons directly, saving additional components. 
This microcontroller controls all sensors and actuators, excepted the internal Li-Po battery recharging logic, which uses a specific chip for safety reasons.
%This choice of a single microcontroller led us to non-trivial design decisions like using low-cost shift registers to drive the many \textsc{led}s.
%Moreover, instead of directly connecting the infrared sensors to an analogical input, we interface them using a comparator that transforms the analogical signal into a binary one with a pulse having a duration proportional to the intensity of the original signal~\cite{wanda}. 
%This design saves \textsc{cpu}-time, reduces power consumption (it allows a shorter infrared pulse than with an analogical design) and allows to use the sensor for communication.

For our specific design, we needed custom made mechanical parts.
To reduce the price, all mechanical parts are in injected plastic, for a total production cost of less than 4\,\$.

Our choice of electronic components implies a production that is either fully automatized or includes some manual operations. 
A final assembly by the user is excluded by the certification required in the case of use by children. 
As a full automatization requires investments that are beyond the possibilities of this project, the current production is based on some manual operations. 
We therefore decided to produce in China, where manual work is less expensive.
%This is a very critical choice in open source hardware projects, and we will discuss this issue in detail in the next section.
We produced until now more than 15,000 robots by batches of 1,000 units, with a cost per robot of 35\,\$.
%Because manual work is expensive in western countries, we produce through subcontractors in China.
%Indeed, while most of the electronics consists of \textsc{smd} components mounted by a robot, some are soldered (sensors, microphone, \textsc{led}s) by hand and the robot is also assembled by hand.
%We streamlined the production of plastic parts (Thymio has only 5 parts while our previous prototype had 11) and opted for injection molding. 
%The plastic hull consists of two main parts screwed together to facilitate the vertical assembly and disassembly of the robot.
%Moreover, when selecting features we took into consideration the complexity of the injection mold.
%The design of the mechanical parts has been aligned with these choicesthat are injected, resulting in all custom parts of Thymio costing less than 4\,\$.
%In these conditions Thymio costs 35\,\$. 
The strict quality control, the production management, the support to the users and the margins for distributors bring to a selling price of 130\,\$.

@@

\subsection{Multi-age and Gender-neutral Feature Set}
\label{sec:multi}

Several design choices, like the variety of sensors, the multiple ways of interacting with the robot, the neutral hull design, the various programming environments, and the possible customization with accessories, contribute to make Thymio accessible to girls and boys of different age groups from kindergarden to university~\cite{Riedo2013}.
These design choices were made and implemented thanks to an important contribution by industrial designers of the University of Art and Design of Lausanne\footnote{\url{http://www.ecal.ch}}.
The white neutral hull is a key element in this set of choices, and it is the opposite than the technical look chosen for the Edison or the LEGO\textsuperscript{\textregistered} robots. 
These last robots are implicitly targeting a group of people interested in technical systems, mostly males, while Thymio is open to both genders and a larger target audience, from 6 years old kids to adults.
%The white hull is initially neutral and children can choose their own color using the \textsc{rgb} \textsc{led}s or attach accessories to the hull.

@@

When designing Thymio, we took care of providing many incentives for the users to learn new things throughout their direct interaction with the robot.
This translates into specific hardware and software choices.

At the hardware level, we render visible the activity of the various robot components by adding a \textsc{led} next to each of them, for a total of 39 \textsc{led}s.
These \textsc{led}s locally color the hull and allow the user to see immediately where and when the robot perceives a change in its environment: proximity of objects, changes in the ground color, temperature, sound, or accelerations.
%\textsc{led}s show, by their light intesity or their color, the measurements of the proximity sensors, the temperature sensor, the microphone, and the accelerometer. 
Some \textsc{led}s display data exchanges, from the infrared remote control receiver or with the microSD card. 
The capacitive buttons give both a visual and an acoustic feedback.
%There is a \textsc{led} next to each proximity sensor that lights up as soon as an object is close enough to be seen, and shines brighter the closer the object gets.
%Similarly, a combination of blue and red \textsc{led}s show the temperature, and the infrared remote control receiver and the microphone both have \textsc{led}s that flash when they detect something.
%On the top of the robot, a circle of 8 \textsc{led}s shows the 3D inclination of the robot thanks to the accelerometer. 
%This circle is also used to reflect some of the behaviors of Thymio.
%Finally, two strong \textsc{rgb} \textsc{led}s color the whole top of the robot.
%In addition to a visual feedback, the capacitive buttons also have an acoustic feedback.
The link between a sensor and its feedback can be turned off when programming the robot to use the \textsc{led}s and the loudspeaker for other purposes.

At the software level, we provide a set of programming environments (see section \ref{sec:aseba}) that enable beginners to discover programming progressively.
First, we teach them the basic rules of programming using a purely visual interface, then they discover the construction of syntax trees by assembling graphical blocks, and finally we provide a full text-based coding environment with advanced debugging tools, such as real-time inspection of the variables of the robot and plotting features, providing a visual way to understand time-related concepts.
%Robotics experts can program Thymio using the \textsc{ros}\footnote{accessible through the \emph{asebaros} bridge, see \url{http://www.ros.org/wiki/asebaros}} environment~\cite{quigley2009ros}.

\subsection{Fast access to robotics behaviors}
\label{sec:behaviors}

>>>>

\author{Francesco Mondada$^{1}$, Michael Bonani$^{2}$, Fanny Riedo$^{2}$, Manon Briod$^{1}$, \\
L\'ea Pereyre$^{1}$, Philippe R\'etornaz$^{1}$ and St\'ephane Magnenat$^{1}$% <-this % stops a space
\thanks{*This research was supported by the Swiss National Center of Competence in Research ``Robotics'' (Thymio robot development and deployment), by GebertRuf Stiftung (design of accessories), by the Swiss NSF project CRAGP2 151543 ``Robotics in schools,'' and by the EU-FP7 project ASSISIbf, no. 601074 (survey on open hardware).
Many thanks to Luc Bergeron and his team at \'ecal.ch for the industrial design of Thymio; Didier Roy, David Sherman and all the INRIA team for their contributions and diffusion in France; Gordana Gerber for the educational material; all the Mobsya team for the effort in production and sale; and Fabian Hahn for the port of Blockly.}% <-this % stops a space``anyone''
\thanks{$^{1}$Francesco Mondada, Manon Briod, L\'ea Pereyre, Philippe R\'etornaz and St\'ephane Magnenat are with the Laboratoire de Syst\`emes Robotiques,
        Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Switzerland.
        {General rule for EPFL emails: \tt\small firstname.lastname@epfl.ch}}%
\thanks{$^{2}$Michael Bonani and Fanny Riedo are with Mobsya Association, Renens, Switzerland.
        {\tt\small michael.bonani@mobsya.org} and {\tt\small fanny.riedo@mobsya.org}}%
}

@@

Mobile robots are valuable tools for education because of both the enthusiasm they raise and the multidisciplinary nature of robotic technology.
Mobile robots give access to a wide range of fields, such as complex mechanics, sensors, wireless transmission, mathematics, and computer science, among others.
As mobile robots sense the environment and take actions based on their perception, they seem to display \emph{intentions of their own}~\cite{Mondada2001}.
%This fascinates the users of the robot and creates feelings of accomplishment and power when these users engage in creative activities with the robots.
This impression of intelligence, the permeating presence of robots in science fiction, and their projected use in our society give a sense of touching the future.
%Children can spend hours looking at a robot interacting with its environment and adults can approach robots as a new hobby that replaces radio and electronics kits.
%The emotional potential of mobile robots and their varied technology makes them a potential ideal tool for education.

Despite their potential as educational tools, robots are still not as widespread in schools as they could be. Among the possible reasons, we believe that the following five play a key role:
\begin{enumerate}
\item Although many research projects are developing innovative and interesting educational robots, few reach sufficient maturity to become distributed and accessible to schools. 
\item A versatile robot performing interesting behaviors is a complex piece of technology and therefore expensive. 
%Indeed, for having an educational value and providing an interesting level of interaction, a robot must embed a wide set of sensors and actuators.
%Existing platforms with these features cost several hundred Euros.
This prevents most schools, which have a limited budget for equipment, from acquiring educational robots.
\item Introducing robotic tools into teaching activities requires investment in time and training for the teachers~\cite{CERI2008}.
%In Europe, despite a trend to increase the use of technological tools in the learning process, teachers are still insufficiently trained and are reluctant to introduce these tools in their teaching activity~\cite{CERI2008}.
Therefore, to be accepted by teachers, robots must be both accessible with minimal effort and accompanied by well-prepared educational material shared among colleagues. 
%Moreover, infrastructure should be provided to allow the sharing of educational material among teachers.
\item Robot construction, use, and programming is often perceived as a boyish activity in our Western society~\cite{leonard2009lego,nourbakhsh2009robot}.
% For instance, several authors report that girls have such a perception of the LEGO Minstorms~\cite{leonard2009lego,nourbakhsh2009robot}.
This strongly limits the potential of robots as general-purpose educational tools, especially in schools.
\item Finally, many teachers are reluctant to follow volatile trends, especially if these are based on purely commercial arguments.
Teachers prefer to invest in stable tools, in contrast to trends in current consumer technology. %, which is often very volatile and based on continuous renewal rather than on durability.
\end{enumerate}

Open source hardware projects can address several of these issues in a different way than closed-source purely commercial products. 
By \textit{open source hardware} we mean, following the definition of the Open Source Hardware Association,\footnote{\url{http://www.oshwa.org/definition/}} ``hardware whose design is made publicly available so that anyone can study, modify, distribute, make, and sell the design or hardware based on that design.''
In this paper, we show that this concept, implemented in the Thymio project through a community of users, developers and manufacturers, brings a strong added value to the robot and to the educational methods.
In addition, we compare our experience with other robotics open source hardware projects not focused on education and highlight challenges and opportunities specific to education.

@@

Many publications present educational robots, from low-cost systems targeting Africa~\cite{Rubenstein2015,Gyebi2015} to extremely sophisticated humanoids~\cite{Hood2015,Mazzoni2016}. 
Among those, only a handful are commercially available, limiting their validation by educational scientists, who are typically not roboticists.
As a result, 90\% of publications about validation of educational results have been based on LEGO\textsuperscript{\textregistered} Mindstorms\textsuperscript{\textregistered}~\cite{benitti2012exploring}, a widely available commercial product.
The latest version, EV3\footnote{\url{http://mindstorms.lego.com}}, is expensive ($\approx$ US\$400) but offers a wide range of possibilities, especially at the mechanical level using LEGO\textsuperscript{\textregistered} bricks and at the software level with its graphical programming environment. 
%NOTE: SM: commented WeDo to save space and streamline story
%At a lower price, LEGO\textsuperscript{\textregistered} proposes the WeDo\textsuperscript{\textregistered} system ($\approx$ 80\,\$).
%The WeDo\textsuperscript{\textregistered} is similar to EV3 but only reads one sensor and controls one actuator and must always be connected to a desktop computer.
%In comparison, the BeeBot robot\footnote{\url{http://www.beebot.org.uk}} is a cheap ($\approx$ 90\,\$) mobile system.
%However, it is not a real robot but rather an automaton: There is no perception of the environment, just a good encoder for precise movements and a simple programming interface based on buttons giving directions for the movement.
Among the other players on the market is the Edison robot\footnote{\url{https://meetedison.com/}}, which is extremely low cost (US\$49) and compatible with LEGO\textsuperscript{\textregistered} bricks. 
The low price has pushed drastic design choices: small size, weak motors, few sensors, communication with the computer by audio jack and standard AAA batteries not included in the price.
These choices strongly limit its possible usage.
%Among widely used robots, none is part of an open source hardware project.

Among the robots available on the market, only a few are open source and are used in schools:
Scribbler2, produced and sold by Parallax\footnote{\url{http://www.parallax.com}} ($\approx$ US\$180), is a 188\,mm large robot, designed to run on the ground and equipped with a few light sensors, one distance sensor, two ground sensors, and few \textsc{led} displays.
It runs on standard AA batteries and has a hacker port for interfacing electronic extensions.
It is programmable with a graphical or a textual code interface.  
The main weakness of Scribbler2 is its limited number of sensors and compatibility with other systems. 
Moreover, there seems to be no active community around its development.
The e-puck~\cite{mondada2009puck} robot targets university-level education.
Well equipped with sensors and actuators, modular and compact, it can be programmed with industry-standard environments.
Several simulators allow running highly complex experiments. 
%There is an active community of users and developers around it.
Its main weakness is its high price ($\approx$ US\$870).
Finch\footnote{\url{http://www.finchrobot.com/}} ($\approx$ US\$99) is a very simple robot that has been designed around a wired connection to the computer. 
This connection reduces electronics requirements, such as batteries or wireless communication, and allows control to be implemented entirely on the computer.
This results in availability of a very broad set of possible programming languages, which is the real force of this robot. 
However, the cable does not allow real autonomy and mobility.
Finally, the mBot\footnote{\url{http://www.makeblock.cc/mbot/}} is a mobile platform based on an Arduino board.
Its electronics is simple and inexpensive and the robot only features a couple of sensors, which allows drastic reduction in its price ($\approx$~US\$75), but also limits the perception possibilities and therefore the span of use.

With respect to these robots, Thymio has a compact size (110\,mm), many interaction possibilities, an affordable price (US\$130), and a large set of sensors.
%Its design makes the best use of the recent cheap components coming from the explosive growth of the mobile devices market. 
To the best of our knowledge, beside Thymio there are no educational products providing a similar integration of sensors and actuators at a lower price.

Teachers, from primary school up to high school, are a primary target user group of the Thymio project.
They decide which tools are used in their class and are key people in the education ecosystem.
For teachers, the motivation to use robotic tools depends on many factors~\cite{chevalier2016}.
Among them, the availability of materials and training plays a key role.
%The BeeBot mentioned previously, because of its radical simplicity, is one of the tools that needs the least training. 
%Moreover its use in learning activities is based on a simple grid on a printed map on the ground, allowing fast and easy integration in existing lessons.
%These are probably two important factors for its success in schools~\cite{2008highfield,2008demichele} and the availability of a rich set of educational material\footnote{\url{https://www.learningplace.com.au/deliver/content.asp?pid=38840} and \\ \url{http://beebots.skola.edu.mt/}}.
%However, the total lack of sensors reduces the BeeBot to an automaton, easy to use but limited to activities implementable by sequences of motions, without any closed-loop control that would allow teaching computational thinking or robotics concepts.
The development of educational material and courses to train teachers requires a huge effort, based on a good mix between robotics and educational skills. 
Moreover, educational material varies from school to school, as requirements are very dependent on local educational programs and languages. 
%For example, in Switzerland, every of the 26 states composing the country has its own school program and its own language among the four national ones. 
%Educational material and the related training must therefore be adapted to each situation.
Even for large companies like LEGO\textsuperscript{\textregistered}, such a development is not sustainable. 
A crowd-sourcing approach may solve this problem; An active community of users can contribute to the development of the material in a distributed manner, adapting the material to the local situation.
LEGO\textsuperscript{\textregistered} itself is moving this direction by promoting communities of users~\cite{Hienerth2014} with a user-producer interaction similar to that of open hardware projects.
An open source community regrouping developers, manufacturers, and end users is therefore a very interesting model to address the distributed development and sharing of educational material and the diffusion of training sessions.
%This model is an extension to education of what is already practiced in projects such as Arduino~\cite{Jamieson2010}.
In this paper, we study a case of implementation of this model.

@@

\begin{figure}
   \centering
	\includegraphics[width=.8\columnwidth]{figures/front_2016}
	\vskip 1.0em
	\includegraphics[width=\columnwidth]{figures/wirelessThymioVecto}
	\caption{The Thymio robot and its main components for the wireless- and the \textsc{usb}-connected versions.}
	\label{fig:schematique}
\end{figure}

We designed the Thymio robot along seven main axes: a low price to address a larger number of users; a feature set that suits both genders and multiple ages from young children to adults; a mechanical design that promotes creativity; a combination of sensors, actuators, and programming features that facilitates learning: a set of ready-to-use programs to quickly access robotic behaviors; an accessible programming environment; and an open source community contributing to design and dissemination.
The result is a miniature differential-wheeled robot suited for use on a desktop (Fig.~\ref{fig:schematique}, top).
The robot is robust enough to be mishandled by children; it can fall from a table without breaking.
It features a translucent white hull and a wide range of sensors and actuators (Fig.~\ref{fig:schematique}, bottom).
The robot has an embedded battery, rechargeable by \textsc{usb}, that provides 3 to 5 hours of power.
More details on the robot and the previous research results can be found in~\cite{RiedoPhD, magnenat2014}.

@@

Price is key in the adoption of robots by schools~\cite{kradolfer2014sociological}.
Thus, the design of Thymio targeted low production costs while including a broad range of functionalities enabling flexibility.
%Therefore the design of Thymio had to target a low production cost, while including a broad range of functionalities to enable flexibility in its use. 
As in this type of robot the main cost comes from electronics and sensors~\cite{thesis_michael}, we focused on low-cost sensors that allow rich interactions with both the environment and the user.

The resulting Thymio robot possesses a large number of simple sensors:
seven horizontal proximity sensors, two ground sensors, a three-axis accelerometer, a thermistor, and a microphone.
Five capacitive touch buttons organized as a direction pad form an intuitive user interface.
Compared to physical buttons, these simplify the plastic hull of the robot and make it more robust.
A remote control receiver provides additional distant buttons.
%A microphone can record sounds and detect claps. 
%Finally, a thermistor measures temperature.
Most of these devices cost less then US\$0.20, the most expensive being the accelerometer with a cost of about US\$0.80, which is an acceptable price given the possibilities it brings to the robot.
%The omnipresence of this type of sensors in smartphones and compact cameras has drastically lowered their prices over the last decade.
We also chose low-cost toy motors and control them in speed (max. 13\,cm/s) measuring the back-electromotive force and therefore removing the need for additional encoders.

%However, choosing the optimal combination of sensors and processor is a difficult problem, as hardware interfaces are limited on a microcontroller and therefore compromises must be made.
We evaluated several microcontrollers and chose the PIC24F from Microchip because it integrates a \textsc{usb} interface and can drive capacitive touch buttons directly, saving additional components. 
This microcontroller controls all sensors and actuators, with the exception of the internal Li-Po battery recharging logic, which uses a specific chip for safety reasons.
%This choice of a single microcontroller led us to non-trivial design decisions like using low-cost shift registers to drive the many \textsc{led}s.
%Moreover, instead of directly connecting the infrared sensors to an analogical input, we interface them using a comparator that transforms the analogical signal into a binary one with a pulse having a duration proportional to the intensity of the original signal~\cite{wanda}. 
%This design saves \textsc{cpu}-time, reduces power consumption (it allows a shorter infrared pulse than with an analogical design) and allows to use the sensor for communication.

For our specific design, we needed custom-made mechanical parts.
To reduce the price, all mechanical parts are injected plastic, for a total production cost of less than US\$4.

Our choice of electronic components implies production that is either fully automatized or includes some manual operations. 
Full assembly is required in order for the robots to be certified for use by children. 
%A final assembly by the user is excluded by the certification required in the case of use by children. 
As a full automatization requires investments that are beyond the possibilities of this project, the current production is based on some manual operations. 
We therefore decided to produce in China, where manual work is less expensive.
%This is a very critical choice in open source hardware projects, and we will discuss this issue in detail in the next section.
We have thus far produced more than 15,000 robots in batches of 1,000 units, with a cost per robot of US\$35.
%Because manual work is expensive in western countries, we produce through subcontractors in China.
%Indeed, while most of the electronics consists of \textsc{smd} components mounted by a robot, some are soldered (sensors, microphone, \textsc{led}s) by hand and the robot is also assembled by hand.
%We streamlined the production of plastic parts (Thymio has only 5 parts while our previous prototype had 11) and opted for injection molding. 
%The plastic hull consists of two main parts screwed together to facilitate the vertical assembly and disassembly of the robot.
%Moreover, when selecting features we took into consideration the complexity of the injection mold.
%The design of the mechanical parts has been aligned with these choicesthat are injected, resulting in all custom parts of Thymio costing less than 4\,\$.
%In these conditions Thymio costs 35\,\$. 
The strict quality control, the production management, the support to the users, and the margins for distributors result in a final selling price of US\$130.

@@

\subsection{Multi-Age and Gender-Neutral Feature Set}
\label{sec:multi}

Several design choices, like the variety of sensors, the multiple ways of interacting with the robot, the neutral hull design, the various programming environments, and the possible customization with accessories, contribute to make Thymio accessible to girls and boys of different age groups from kindergarten to university~\cite{Riedo2013}.
These design choices were made and implemented thanks to an important contribution by industrial designers of the University of Art and Design of Lausanne.\footnote{\url{http://www.ecal.ch}}
The white neutral hull is a key element in this set of choices, and it is the opposite of the technical look chosen for the Edison or the LEGO\textsuperscript{\textregistered} robots. 
These latter two robots implicitly target a group of people interested in technical systems, mostly males, while Thymio is open to both genders and a larger target audience.
%The white hull is initially neutral and children can choose their own color using the \textsc{rgb} \textsc{led}s or attach accessories to the hull.

@@

When designing Thymio, we took care to provide many incentives for the users to learn new things throughout their direct interaction with the robot.
This translates into specific hardware and software choices.

At the hardware level, we render visible the activity of the various robot components by adding an \textsc{led} next to each of them, for a total of 39 \textsc{led}s.
These \textsc{led}s locally color the hull and allow the user to see immediately where and when the robot perceives a change in its environment: proximity of objects, changes in the ground color, temperature, sound, or accelerations.
%\textsc{led}s show, by their light intesity or their color, the measurements of the proximity sensors, the temperature sensor, the microphone, and the accelerometer. 
Some \textsc{led}s display data exchanges from the infrared remote control receiver or with the microSD card. 
The capacitive buttons give both visual and acoustic feedback.
%There is a \textsc{led} next to each proximity sensor that lights up as soon as an object is close enough to be seen, and shines brighter the closer the object gets.
%Similarly, a combination of blue and red \textsc{led}s show the temperature, and the infrared remote control receiver and the microphone both have \textsc{led}s that flash when they detect something.
%On the top of the robot, a circle of 8 \textsc{led}s shows the 3D inclination of the robot thanks to the accelerometer. 
%This circle is also used to reflect some of the behaviors of Thymio.
%Finally, two strong \textsc{rgb} \textsc{led}s color the whole top of the robot.
%In addition to a visual feedback, the capacitive buttons also have an acoustic feedback.
The link between a sensor and its feedback can be turned off when programming the robot so that the \textsc{led}s and loudspeaker can be used for other purposes.

At the software level, we provide a set of programming environments (see the Programming Environment section below) that enable beginners to discover programming progressively.
First, we teach them the basic rules of programming using a purely visual interface, then they discover the construction of syntax trees by assembling graphical blocks, and finally we provide a full text-based coding environment with advanced debugging tools, such as real-time inspection of the variables of the robot and plotting features, providing a visual way to understand time-related concepts.
%Robotics experts can program Thymio using the \textsc{ros}\footnote{accessible through the \emph{asebaros} bridge, see \url{http://www.ros.org/wiki/asebaros}} environment~\cite{quigley2009ros}.

\subsection{Fast Access to Robotics Behaviors}
\label{sec:behaviors}
